---
title: "Data viz basics with R and ggplot2"
author: "Quentin D. Read"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This is a 

Students should already have a beginner-level knowledge of R, including basic knowledge of functions and syntax, and awareness of how data frames in R work.

## Learning objectives

At the end of this course, you will know ...

- what the "grammar of graphics" is and how **ggplot2** uses it
- how to map a variable in a data frame to a graphical element in a plot using `aes`
- how to use different `geom`s to make scatterplots, boxplots, and histograms
- how to add trendlines to a plot
- how to make plots for different subsets of your data using `facet`s
- how to change the style and appearance of your plots

## What is the grammar of graphics?

## How does a ggplot work?

- data
- aesthetic mappings
- one or more geoms (layers)
- scales
- themes

## The data

We will be using three fun datasets from [Kaggle](https://www.kaggle.com), a data science competition site where users upload public domain datasets. Click on each link if you want to learn more about each dataset, including descriptions of each column.

- [Wine reviews posted on winemag.com in 2017](https://www.kaggle.com/datasets/zynicide/wine-reviews). *I've given you only the first 1000 rows of this dataset.*
- [World Happiness Report 2015](https://www.kaggle.com/datasets/unsdsn/world-happiness)
- [Nutritional values of 77 brands of cereal](https://www.kaggle.com/datasets/crawford/80-cereals)

# The code

## Loading packages and data

We will use only the **ggplot2** package in this tutorial. Use the `read.csv()` function from base R to read in each of the three datasets.

```{r}
library(ggplot2)
```

